Coalition Formation Algorithm of Prosumers in a Smart Grid Environment

### 2014-09-15 Article.tex
update Nico

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Abstract
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
In a smart grid environment, we study coalition formation of prosumers that aim at selling energy to the grid. It is paramount for the grid operation that the energy producers are able to sustain the grid demand in terms of stability and minimum production requirement in order to enter the energy market. We design an algorithm that seeks to form coalitions that will meet both of these requirements:  a minimum energy level for the coalitions and a steady production level which leads to finding uncorrelated source of energy  to form a coalition. We propose an algorithm that uses graph tools such as correlation graphs or clique percolation to form coalitions that meet such complex constraints. We validate the algorithm against a random procedure and show that, it not only performs better in term of social welfare for the power grid, but also that it is more robust against unforeseen production variations due to changing weather conditions for instance. 
\end{abstract}

@@

In this paper, we focus on how to statistically improve the production stability by carefully forming coalitions of prosumers that have greater probabilities of staying in acceptable range of production. By using prediction techniques, we form coalitions that meet the contract values ranges proposed by the grid, enabling it to schedule its production accordingly. Meeting contract values for the prosumers is especially relevant in power trade conditions, where energy is traded based on day-ahead predictions. In this context, participants should try to minimize their prediction errors in order to maintain a stable state for the grid. They may even occur some penalties if the productions deviate from the initial contract values. However, it remains pertinent for both side to maintain the production as stable as possible : for the prosumer, the stability (with renewable energy source) will impact its net gain, and for the power grid, it will helps to maintain its reliability.

@@

$\epsilon$-graphs, consists in filtering edges based on their weight, only keeping edges whose weights are less than $ \epsilon $. In \cite{Garas2008, Onnela2004}, the authors studied the topological properties (average clustering, connectivity, relative number of cliques) of the correlation graph against those of growing random graphs, depending on the threshold $ \epsilon $. However, there is not well defined method to select of right tradeoff $ \epsilon $ as function of the network topology.

@@

At this point, modeling agents consists in fixing a few parameters (most of the time drawn from random distributions) such as the geographical position, the number of wind turbines, PV, appliances, the temperature of confort, and so on. The objective is that, depending on the weather of his zone, the DER and appliances he owns, and the way he decides to heat his home, a given prosumer is able to compute his production/consumption at any time. We thus used simple and known power curve models for linking weather variables to output power, such that a wind turbine (or a PV for instance) takes a wind speed (a solar irradiance) as input and gives an instantaneous output power \cite{Kota2011} \cite{windturbinemodel}. Any agent is thus able to compute at any time how much he produces and consumes.

\subsection{Coalition formation model}\label{subsec:Coalition}
More formally, we denote by $ \mathcal{A} = \{ a_{1},...a_{N} \} $ the set of agents (the prosumers) and $ P_{i}(t) $ the instantaneous power value at agent i at instant t (its instantaneous production minus his instantaneous consumption, meaning that $ P_{i}(t) $ represents agent i available power at instant t). During the simulation (from $t_{0} $ to $ t_{K} $), all agents record their values with an hour time interval. As expected (because we introduced them), the time-series exhibit high seasonal patterns, but completely different from agent to agent because they depend both of their energetic mix and habits of a given prosumer. Nevertheless, these macro variations can hide the interesting information in the correlation coefficients and twist the rest of our process. Thereby, we remove these seasonal effects for all agents (see second block of figure \ref{Fig1}). We denote by $ \mathcal{T}_{i} = \{ P_{i}(t_{0}),...,P_{i}(t_{K}) \} $ the resulting time-series for  agent i. For any agent i, we note $ \mathcal{P}_{i} $ the probability distribution drawn from $ \mathcal{T}_{i} $ and by $ P_{i} $ a random variable following $ \mathcal{P}_{i} $.

We now extend these notations to any coalition $ S \subset \mathcal{A} $ : 
\begin{itemize}
\item $ P_{S}(t) = \sum_{i \in S} P_{i}(t) $
\item $ \mathcal{T}_{S} = \{ P_{S}(t_{0}),...,P_{S}(t_{K}) \} $ with $ \mathcal{P}_{S} $ its probability distribution
\item A coalition S has the possibility to announce a contract value $ P_{S}^{CRCT} $ on the market.
\end{itemize}

In this paper, we consider that coalition announcements are constrained by the grid. Namely, the grid has the possibility of fixing two parameters :

\begin{itemize}
\item the \textbf{reliability} (denoted by $ \phi \in [0,1] $)
\item the \textbf{minimum contract value} ( $ P_{\phi}^{MIN} $). 
\end{itemize}

The reliability stipulates that for any coalition S willing to join the market, the probability of S value (at any instant t) being below its contract value should at most be $ \phi $. That is, $ \forall S,\ \forall t,\ Pr[P_{S}(t) \leq P_{S}^{CRCT} ] \leq \phi $ (for consistency we restrict $ \phi $ to small values ( $ \phi << 0.5 $)). The minimum contract value under $ \phi $ states that only coalitions with higher contract values ($ P_{S}^{CRCT} \geq P_{\phi}^{MIN} $) will be accepted.

At this point, if a coalition S wishes to join the market, it has to choose a contract value that fulfill the two conditions. For simplification, we consider that coalitions will always apply the same economically consistent strategy of announcing the highest possible contract value that obeys the reliability rule (a value we denote by $ P_{\phi}(S) $). If $ P_{\phi}(S) \geq P_{\phi}^{MIN} $, meaning that it also obeys the minimum contract value rule, then S announces this value on the market : $ P_{S}^{CRCT} = P_{\phi}(S) $. Otherwise, coalition S is not able to enter.
Basically, a coalition S is valid if and only if :
\begin{equation}
\left\{ \begin{array}{lll}
		\forall t,\ Pr[ P_{S}(t) \leq P_{\phi}(S)] \leq \phi\ \textit{{\scriptsize (reliability rule)}} \\
		and\ P_{\phi}(S) \geq P_{\phi}^{MIN}\ \textit{{\scriptsize (min value rule)}}

@@

\begin{equation}
\tau_{i}(S) = \dfrac{\mathcal{U}_{\phi,\ P_{\phi}^{MIN}}(S) - \mathcal{U}_{\phi,\ P_{\phi}^{MIN}}(S-\{i\}) }{\mathcal{U}_{\phi,\ P_{\phi}^{MIN}}(S)}
\label{tau}
\end{equation}

@@

At this point, we have three degrees of freedom : the reliability ($ \phi $), the required power to enter the market ($P_{\phi}^{MIN}$), and the prunning parameter ($\epsilon$). For convenience, we introduce the number of desired coalitions ($ N_{COAL} $) and we fix $\epsilon $ as the smallest value such that the $ \epsilon $-graph contains at least $ N_{COAL} $ non overlaping cliques. The overall utility depends now on the grid requirements and on the number of desired coalitions.

@@

\begin{figure}
 \centering
  \subcaptionbox{$ (\phi, P_{\phi}^{MIN}) $ space\label{fig2:a}}{\includegraphics[width=1.6in]{./figure1/util}}\hspace{1em}%
  \subcaptionbox{$ (N_{COAL}, P_{\phi}^{MIN}) $ space\label{fig2:b}}{\includegraphics[width=1.6in]{./figure6/util2}}
\caption{Utility function in parameters space : In $ (\phi, P_{\phi}^{MIN}) $ when $ N_{COAL} $ is fixed (fig \ref{fig2:a}), and in $ (N_{COAL}, P_{\phi}^{MIN}) $ for a fixed $ \phi $ (fig \ref{fig2:b})}
\label{Fig2}
\end{figure}

As visible on figure \ref{fig2:a}, the parameters $\phi$ and $P_{\phi}^{MIN}$ shape the utility function such that, if $ \phi $ is close to zero, the reliability requirement is very high and only small values of $ P_{\phi}^{MIN}$ could potentially lead to valid coalitions (and positive utility values). Conversely, the higher $\phi$, the less constraints are imposed to the coalitions and valid ones can arise for a larger spectrum of $ P_{\phi}^{MIN}$. Obviously, the highest utility values are found for high $ \phi $, because coalitions are able then to announce higher contract values, yielding higher utilities.

In the following, we fix the reliability to a given empirical value ($\phi = Cste $) and observe how the coalitions evolves for different values of $P_{\phi}^{MIN}$. Note that the opposite is also possible although a little less intuitive, but not shown here because of space limitations and because it leads to similar conclusions.

On figure \ref{fig2:b}, we can see that when the grid constraints are fixed ($ \phi = 0.08 $ and $ P_{\phi}^{MIN} = 2 $ for instance) and only the number of coalitions varies, there is an increase in social welfare (sum of coalitions utilities) up to a maximum point (for $ N_{COAL} = 3 $ here) before a decrease. The reason is that increasing the number of coalitions allows more coalitions to reach stability and enter the market, but there is a point where nodes bringing stability are not sufficient inside the coalitions to make them pass the grid requirements, and some coalitions start to fail with zero utility. Moreover, reckon that increasing $ N_{COAL} $ means also increasing $ \epsilon $, leading to denser graphs where information is flooded, meaning that the algorithm performances also decrease.
Naturally, it is also visible that small values of $P_{\phi}^{MIN}$ lead to higher utilities because coalitions are able to announce higher contract values.

\begin{center}
\begin{figure}[b]
  \includegraphics[scale=0.8]{./figure5/coals}
  \caption{Evolution of the coalitions for different values of $ P_{\phi}^{MIN} $ for clique percolation (blue dots) and random process (red dots). The diameters of the dots are proportional to the corresponding coalition sizes.}
  \label{Fig3}
\end{figure}
\end{center}

For our 200 prosumers, we suppose that the grid fixes a reliability $ \phi = 0.1 $ (meaning that coalitions should produce more than their contract values at least $ 90 \% $ of the time). We also fix the number of coalitions to 10 and have a look at how these coalitions evolve as the grid changes the $ P_{\phi}^{MIN} $ requirement. As a comparison, we use a completely random algorithm that only asks for a number of coalitions and partitions the agents in a random fashion. Figure \ref{Fig3} shows this evolution for our algorithm (blue dots) and for the random process (red dots). The diameter of a dot is proportional to the number of agents in the coalition and the higher the dot is, the higher its utility. The $ P_{\phi}^{MIN} $ values of the x axis are expressed in tenth of MW for readability and the hatched zone corresponds to the "under-requirement space", meaning that whenever a coalition is in this zone, it has a null utility.

Looking at figure \ref{Fig3}, we can see first that our algorithm seems to make generally a better job at finding high valued coalitions, but also that the results seem far more robust against how the grid positions its requirements. The blue dots allowed to enter the market seem indeed to outnumber the red dots, especially when the grid requirements are neither too low or too high ($ P_{\phi}^{MIN} = 5 $ in figure \ref{Fig3}).

In more details, figure \ref{fig4:a} presents the evolution of social welfare as the number of coalitions increases (all other parameters are kept constant) for random (red curve), clique percolation (blue curve), and correlated coalitions (green curves) that stands for a worst case scenario. As for figure \ref{fig4:b}, it shows the percentage of coalitions able to enter the market for different values of $ P_{\phi}^{MIN} $. For consistency, we average the results of both plots of the random procedure over 100 realizations, and the errorbars in the plots stands for the standard deviations of the results.

@@

When the grid requirements vary, for very low $ P_{\phi}^{MIN} $, all coalitions for all algorithms are able to enter the market, yielding an acceptance percentage of $ 100 \% $. But as $ P_{\phi}^{MIN} $ increases, we see the percentage of the correlated coalitions crashing down very quickly. After a few increases in $ P_{\phi}^{MIN} $, the percentage of the random procedure starts droping while it stays almost constant for our algorithm. For $ P_{\phi}^{MIN} = 8 $, we see that only a little more than half of the coalitions for the random case are able to enter while approximately $ 85 \%$ of them enters for the clique percolation algorithm. Finally, when the grid requirement becomes too high, the acceptance percentage of our algorithm tends slowly to zero (not shown in this plot for readability).

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Abstract
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
In a smart grid environment, we study coalition formation of prosumers that aim at selling energy to the grid. It is paramount for the grid operation that the energy producers are able to sustain the grid demand in terms of stability and minimum production requirement in order to enter the energy market. We design an algorithm that seeks to form coalitions that will meet both of these requirements:  a minimum energy level for the coalitions and a steady production level which leads to finding uncorrelated sources of energy  to form a coalition. We propose an algorithm that uses graph tools such as correlation graphs or clique percolation to form coalitions that meet such complex constraints. We validate the algorithm against a random procedure and show that, it not only performs better in term of social welfare for the power grid, but also that it is more robust against unforeseen production variations due to changing weather conditions for instance. 
\end{abstract}

@@

In this paper, we focus on how to statistically improve the production stability by carefully forming coalitions of prosumers that have greater probabilities of staying in acceptable range of production. By using prediction techniques, we form coalitions that meet the contract values ranges proposed by the grid, enabling it to schedule its production accordingly. Meeting contract values for the prosumers is especially relevant in power trade conditions, where energy is traded based on day-ahead predictions. In this context, participants should try to minimize their prediction errors in order to maintain a stable state for the grid. They may even occur some penalties if the productions deviate from the initial contract values. However, it remains pertinent for both side to maintain the production as stable as possible : for the prosumer, the stability (with renewable energy sources) will impact its net gain, and for the power grid, it will helps to maintain its reliability.

@@

$\epsilon$-graphs, consists in filtering edges based on their weight, only keeping edges whose weights are less than $ \epsilon $. In \cite{Garas2008, Onnela2004}, the authors studied the topological properties (average clustering, connectivity, relative number of cliques) of the correlation graph against those of growing random graphs, depending on the threshold $ \epsilon $. However, there is no well defined method to select of right tradeoff $ \epsilon $ as function of the network topology.

@@

At this point, modeling agents consists in fixing a few parameters (most of the time drawn from random distributions) such as the geographical position, the number of wind turbines, PV, appliances, the temperature of confort, and so on. The objective is that, depending on the weather of his zone, the DER and appliances he owns, and the way he decides to heat his home, a given prosumer is able to compute his production/consumption at any time.

We thus used existing models enabling conversion between weather variables and output powers (see power curves in \cite{Kota2011} \cite{windturbinemodel}). \cite{Dans2007} provides also a convenient way of using cloudiness traces as realistic degradation factors in a \textit{"clear blue sky"} solar model, which enables us to rebuild realistic solar irradiance traces for stations that do not provide such information directly. Any agent is thus able to compute at any time how much he produces and consumes.

\subsection{Coalition formation model}\label{subsec:Coalition}
More formally, we denote by $ \mathcal{A} = \{ a_{1},...a_{N} \} $ the set of agents (the prosumers) and $ P_{i}(t) $ the instantaneous power value at agent i at instant t (its instantaneous production minus his instantaneous consumption, meaning that $ P_{i}(t) $ represents agent i available power at instant t). During the simulation (from $t_{0} $ to $ t_{K} $), all agents record their values with an hour time interval. As expected (because we introduced them), the time-series exhibit high seasonal patterns, but completely different from agent to agent because they depend both of their energetic mix and habits of a given prosumer. Nevertheless, these macro variations can hide the interesting information in the correlation coefficients and twist the rest of our process. Thereby, we remove these seasonal effects for all agents (see second block of figure \ref{Fig1}). We denote by $ \mathcal{T}_{i} = \{ P_{i}(t_{0}),...,P_{i}(t_{K}) \} $ the resulting time-series for  agent i, and by $ \mathcal{P}_{i} $ the corresponding probability distribution.

We now extend these notations to any coalition $ S \subset \mathcal{A} $ : 
\begin{itemize}
\item $ P_{S}(t) = \sum_{i \in S} P_{i}(t) $
\item $ \mathcal{T}_{S} = \{ P_{S}(t_{0}),...,P_{S}(t_{K}) \} $ with $ \mathcal{P}_{S} $ the corresponding probability distribution
\item A coalition S has the possibility to announce a contract value $ P_{S}^{CRCT} $ on the market.
\end{itemize}

Due to the volatile nature of DER productions or user consumptions, the available power of a coalition will oscillates around $ P_{S}^{CRCT} $. It will be the responsability of the coalition, that is the information system maintaining the coalition, to compensate for deviations. In this paper, we consider then that the smaller these deviations, the better. Based on the $ \mathcal{T}_{S} $ distribution, the probability of S value (at any instant t) being below its contract value $ Pr[P_{S}(t) \leq P_{S}^{CRCT} ] $ appears therefore as a quantity of interest that we would want to keep as low as possible.

In this model, we consider that only the coalitions that have this probability value sufficiently low are acceptable. That this, we denote by $ \phi \in [0,1] $ the threshold imposed to the coalitions over this quantity, and we call it the \textbf{reliability} in the following. The reliability stipulates then that for any coalition S willing to join the market, the probability of S value (at any instant t) being below its contract value should at most be $ \phi $. That is, $ \forall S,\ \forall t,\ Pr[P_{S}(t) \leq P_{S}^{CRCT} ] \leq \phi $ ($ \phi $ being restricted to small values for consistency).

We then consider another constraint denoted by $ P^{MIN} $, as the \textbf{minimum contract value}, which states that only coalitions with higher contract values ($ P_{S}^{CRCT} \geq P^{MIN} $) will be accepted.

At this point, if a coalition S wishes to join the market, it has to choose a contract value that fulfill the two conditions. For simplification, we consider that coalitions will always apply the same economically consistent strategy of announcing the highest possible contract value that obeys the reliability rule (a value we denote by $ P_{\phi}(S) $). If $ P_{\phi}(S) \geq P^{MIN} $, meaning that it also obeys the minimum contract value rule, then S announces this value on the market : $ P_{S}^{CRCT} = P_{\phi}(S) $. Otherwise, coalition S is not able to enter.
Basically, a coalition S is valid if and only if :
\begin{equation}
\left\{ \begin{array}{lll}
		\forall t,\ Pr[ P_{S}(t) \leq P_{\phi}(S)] \leq \phi\ \textit{{\scriptsize (reliability rule)}} \\
		and\ P_{\phi}(S) \geq P^{MIN}\ \textit{{\scriptsize (min value rule)}}

@@

\begin{equation}
\tau_{i}(S) = \dfrac{\mathcal{U}_{\phi,\ P^{MIN}}(S) - \mathcal{U}_{\phi,\ P^{MIN}}(S-\{i\}) }{\mathcal{U}_{\phi,\ P^{MIN}}(S)}
\label{tau}
\end{equation}

@@

At this point, we have three degrees of freedom : the reliability ($ \phi $), the required power to enter the market ($P^{MIN}$), and the prunning parameter ($\epsilon$). For convenience, we introduce the number of desired coalitions ($ N_{COAL} $) and we fix $\epsilon $ as the smallest value such that the $ \epsilon $-graph contains at least $ N_{COAL} $ non overlaping cliques. The overall utility depends now on the grid requirements and on the number of desired coalitions.

@@

\begin{figure}
 \centering
  \subcaptionbox{$ (\phi, P^{MIN}) $ space\label{fig2:a}}{\includegraphics[width=1.6in]{./figure1/util}}\hspace{1em}%
  \subcaptionbox{$ (N_{COAL}, P^{MIN}) $ space\label{fig2:b}}{\includegraphics[width=1.6in]{./figure6/util2}}
\caption{Utility function in parameters space : In $ (\phi, P^{MIN}) $ when $ N_{COAL} $ is fixed (fig \ref{fig2:a}), and in $ (N_{COAL}, P^{MIN}) $ for a fixed $ \phi $ (fig \ref{fig2:b})}
\label{Fig2}
\end{figure}

As visible on figure \ref{fig2:a}, the parameters $\phi$ and $P^{MIN}$ shape the utility function such that, if $ \phi $ is close to zero, the reliability requirement is very high and only small values of $ P^{MIN}$ could potentially lead to valid coalitions (and positive utility values). Conversely, the higher $\phi$, the less constraints are imposed to the coalitions and valid ones can arise for a larger spectrum of $ P^{MIN}$. Obviously, the highest utility values are found for high $ \phi $, because coalitions are able then to announce higher contract values, yielding higher utilities.

In the following, we fix the reliability to a given empirical value ($\phi = Cste $) and observe how the coalitions evolves for different values of $P^{MIN}$. Note that the opposite is also possible although a little less intuitive, but not shown here because of space limitations and because it leads to similar conclusions.

On figure \ref{fig2:b}, we can see that when the grid constraints are fixed ($ \phi = 0.08 $ and $ P^{MIN} = 2 $ for instance) and only the number of coalitions varies, there is an increase in social welfare (sum of coalitions utilities) up to a maximum point (for $ N_{COAL} = 3 $ here) before a decrease. The reason is that increasing the number of coalitions allows more coalitions to reach stability and enter the market, but there is a point where nodes bringing stability are not sufficient inside the coalitions to make them pass the grid requirements, and some coalitions start to fail with zero utility. Moreover, reckon that increasing $ N_{COAL} $ means also increasing $ \epsilon $, leading to denser graphs where information is flooded, meaning that the algorithm performances also decrease.
Naturally, it is also visible that small values of $P^{MIN}$ lead to higher utilities because coalitions are able to announce higher contract values.

\begin{center}
\begin{figure}[b]
  \includegraphics[scale=0.8]{./figure5/coals}
  \caption{Evolution of the coalitions for different values of $ P^{MIN} $ for clique percolation (blue dots) and random process (red dots). The diameters of the dots are proportional to the corresponding coalition sizes.}
  \label{Fig3}
\end{figure}
\end{center}

For our 200 prosumers, we suppose that the grid fixes a reliability $ \phi = 0.1 $ (meaning that coalitions should produce more than their contract values at least $ 90 \% $ of the time). We also fix the number of coalitions to 10 and have a look at how these coalitions evolve as the grid changes the $ P^{MIN} $ requirement. As a comparison, we use a completely random algorithm that only asks for a number of coalitions and partitions the agents in a random fashion. Figure \ref{Fig3} shows this evolution for our algorithm (blue dots) and for the random process (red dots). The diameter of a dot is proportional to the number of agents in the coalition and the higher the dot is, the higher its utility. The $ P^{MIN} $ values of the x axis are expressed in tenth of MW for readability and the hatched zone corresponds to the "under-requirement space", meaning that whenever a coalition is in this zone, it has a null utility.

Looking at figure \ref{Fig3}, we can see first that our algorithm seems to make generally a better job at finding high valued coalitions, but also that the results seem far more robust against how the grid positions its requirements. The blue dots allowed to enter the market seem indeed to outnumber the red dots, especially when the grid requirements are neither too low or too high ($ P^{MIN} = 5 $ in figure \ref{Fig3}).

In more details, figure \ref{fig4:a} presents the evolution of social welfare as the number of coalitions increases (all other parameters are kept constant) for random (red curve), clique percolation (blue curve), and correlated coalitions (green curves) that stands for a worst case scenario. As for figure \ref{fig4:b}, it shows the percentage of coalitions able to enter the market for different values of $ P^{MIN} $. For consistency, we average the results of both plots of the random procedure over 100 realizations, and the errorbars in the plots stands for the standard deviations of the results.

@@

When the grid requirements vary, for very low $ P^{MIN} $, all coalitions for all algorithms are able to enter the market, yielding an acceptance percentage of $ 100 \% $. But as $ P^{MIN} $ increases, we see the percentage of the correlated coalitions crashing down very quickly. After a few increases in $ P^{MIN} $, the percentage of the random procedure starts droping while it stays almost constant for our algorithm. For $ P^{MIN} = 8 $, we see that only a little more than half of the coalitions for the random case are able to enter while approximately $ 85 \%$ of them enters for the clique percolation algorithm. Finally, when the grid requirement becomes too high, the acceptance percentage of our algorithm tends slowly to zero (not shown in this plot for readability).
